---
title: "Inferring mutational signatures with uncertainty using hdpExtra"
author: "Víctor Velasco-Pardo"
date: "March 2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantifying uncertainty in mutational signare analyses with hdpExtra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up using the hdp package

## Matrix of mutation counts

In general, we are presented with a matrix of mutation counts, $M$, where rows
index patients and columns index mutation categories. The package hdp allows us 
to set up a hierarchy of Dirichlet Processes to model such data, with an 
additional level in the hierarchy to account for grouping structure (such as
different cancer types) if needed. 

To illustrate this, we use the the dataset of 21 breast cancers by Nik-Zainal et
al. (2012). Using Non-negative matrix factorisation, Nik-Zainal et al. found 
5 signatures in that dataset. Here, we aim to reproduce that analysis using the 
hdp package of Roberts (2018) and introducing a novel post-processing approach.
We downloaded the data from the FTP server and constructed the mutational counts
matrix using SigProfilerMatrixGenerator.

```{r}
library(hdp)
library(hdpExtra)
data("breast_counts")
dim(breast_counts)
tail(sort(rowSums(breast_counts)))
```
Note that one of the 21 patients, "PD4120a", is hyper-mutated. While other authors
have suggested to "down-sample" the mutation counts of hyper-mutated patients,
we strongly advise users against doing so (explain). 

## Initialising the HDPM model

We use the <tt>hdp</tt> package to initialise an objects necessary for making 
inference with the Hierarchical Dirichlet Process. 

```{r}
# initialise HDP
hdp_mut <- hdp::hdp_init(ppindex = c(0, rep(1, each = nrow(breast_counts))), # index of parental nodes
                         cpindex = c(1, rep(2, each = nrow(breast_counts))), # index of concentration param
                         hh=rep(1, 96), # prior is uniform over the 96 mutation categories
                         alphaa=c(1.0,1.0), # shape hyperparams for five different CPs
                         alphab=c(0.1,1.0)) # rate hyperparams for five different CPs

# add data to leaf nodes (one per cancer sample, in row order of mut_count)
hdp_mut <- hdp::hdp_setdata(hdp_mut,
                            dpindex = 2:numdp(hdp_mut), # index of nodes to add data to
                            breast_counts) # input data (mutation counts, sample rows match up with specified dpindex)
hdp_mut
```

### Mathematical setting

The model we have initialised can be described mathematically as 
$$
\begin{align}
G_0 &\sim \text{DP}(\gamma, H) \\
G_j &\sim \text{DP}(\alpha, G_0), \quad j = 1,\dots,J,
\end{align}
$$
where $H$ is the base measure, $\gamma$ and $\alpha$ are concentration parameters,
and $j$ indexes patients in the dataset. $G_0$ is a the centering distribution
around mutational signatures, and it is itself a random object, and $G_j$ is the
distribution around mutational signatures characterising the $j$th patient.
Note that there is one concentration parameter at the top of the hierarchy, $\gamma$,
and one concentration parameter at the patient level, $\alpha$. Unlike other authors,
we discourage users from using one concentration parameter per patient (that is,
we discourage users from setting $G_j \sim \text{DP}(\alpha_j,G_0)$. 

### Prior on the concentration parameters

Concentration parameters, $\gamma$ and $\alpha$, determine how the number of
mutational signatures grows with the sample size, in the sample and in individual
patients respectively. Because this number is not know in advance, we set a
vague (hyper)prior on both hyperparameters. Following the original work of Teh et al.,
we set the following vague gamma priors:
$$
\begin{align}
\gamma &\sim \Gamma(1, 0.1) \\
\alpha &\sim \Gamma(1, 1)
\end{align}
$$


# Inferring signatures with hdpExtra

Now we run two MCMC chains to draw samples from the joint posterior distribution.
First, we initialise each of the chains using the method dp_activate in hdp. Then, 
we use the method hdpExtra_posterior in hdp, which is a modified version of hdp's
hdp_posterior. Unlike hdp_posterior, hdpExtra_posterior stores the sampled vector
of cluster labels that characterises a partition of the mutations, hence allowing
us to summarise the posterior distribution. Additionally, hdpExtra_posterior stores
the concentration parameters and the number of clusters across iterations of the 
sampler, allowing us to check for convergence (see below).
We set the initial number of cluster to a number much higher than the one we expect.
In this example, we set the initial number of cluster to 100.

```{r}
chlist <- vector("list", 2)
for (i in 1:2){

  # activate DPs, 10 initial components
  hdp_activated <- hdp::dp_activate(hdp_mut, 1:numdp(hdp_mut), initcc=100, seed=i*200)

  chlist[[i]] <- hdpExtra_posterior(hdp_activated,
                               burnin=100,
                               n=10,
                               space=10,
                               cpiter=3,
                               seed=i*1e3)
}
```

Now chlist contains a list of objects of class HdpExtraChain. In turn, each such
object stores, for each iteration of the sampler, the aforementioned partitions, 
the mutational signatures associated with that partition, and an object of class
hdpSampleChain, for compatibility with hdp, amongst other attributes. The samples
from the joint posterior over the partition and the signatures allow us to characterise
the uncertainty around the reported solution. 

Having obtained draws from the posterior distribution across several chains, we 
can combine the output of the chains in an object of class HdpExtraChainMulti
```{r}
chlist <- HdpExtraChainMulti(chlist)
```

# Assessig convergence

# Plotting mutational signatures

# Compatibility with hdp

The output of hdpExtra_posterior can also be used for the post-processing tools
of hdp. 
